NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-06 Manuscript/manuscriptPS.tex
Style edits to methods, electrometer.

\subsection{Comparison of ion binding to negatively charged lipid bilayers 
between simulations and experiments using the electrometer concept}

The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
can be used to measure the ion binding affinity because they 
decrease proportionally to the amound of bound positive
charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
This molecular electrometer concept is especially useful for 
the comparison between simulations and experiments because
the headgroup order parameters can be directly calculated from 
simulations and compared to the experimental data with varying cation
concentrations~\cite{catte16}. The headgroup order parameters
of negatively charged PS and PG lipids exhibit systemic, but less
characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized by measuring the PC headgroup order parameters from 
mixed bilayers~\cite{roux86,roux90,roux91}, see section \ref{electrometerFORmixtures} in the supplementary information.

Before using the PC headgroup order parameters to quantify the ion binding
affinity, it is important to quantify their response to the known amount of bound charge~\cite{catte16,melcr18}.
This can be done using the experimental data from the mixtures of
monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
see section \ref{electrometerCALIBRATION} in the supplementary information.
In this work, we also quantify the response of PC headgroup order parameters
to the negatively charged PS headgroups, which also follows the electrometer
concept in the experiments~\cite{scherer87},
see section \ref{electrometerFORmixtures} in the supplementary information.

In the experimental $^2$H NMR literature data used in this work~\cite{browning80,roux90},
the lipids were first soluted to the buffer and then centrifuged to a pellet which
was used in the measurements. Such samples have lower lipid concentration
(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
Larger multilamellar repeat distances are expected in the samples with lower lipid
concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
However, the PS headgroup order parameters measured from gravimetric sample in this work
are in good agreement with the results from centrifuged sample in the literature \cite{browning80} (Fig. \ref{HGorderParameters}).
Furthermore, the equlibrium repeat distance rapidly decreases with the addtion of monovalent
salts and is close to the simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
in the used simulations and reference experiments.

Two different definitions for the salt concentrations have been used when
electrometer concept is applied to study ion binding affinity.
The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. The used definition has only a marginal effect
to the results is simulations with realistic ion binding affinity
(section \ref{concentrationDEFsection} in the supplementary information).

>>>>

\subsection{Comparison of ion binding to negatively charged lipid bilayers 
between simulations and experiments using the molecular electrometer}

As the order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
they can be used to measure the ion binding affinity.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup order parameters at varying cation
concentrations can be directly calculated from 
simulations and compared to the corresponding experimental data~\cite{catte16}. The headgroup order parameters
of negatively charged PS and PG lipids also exhibit systematic, but less
understood dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized by measuring the PC headgroup order parameters from 
mixed bilayers~\cite{roux86,roux90,roux91}, see section \ref{electrometerFORmixtures} in the supplementary information.

Before using the PC headgroup order parameters to quantify the ion binding
affinity, it is important to quantify their response to a known amount of bound charge~\cite{catte16,melcr18}.
This can be done using experimental data from mixtures of
monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
see section \ref{electrometerCALIBRATION} in the supplementary information.
In this work, we also quantify the response of PC headgroup order parameters
to the negatively charged PS headgroups, which also follows the molecular electrometer
concept in experiments~\cite{scherer87},
see section \ref{electrometerFORmixtures} in the supplementary information.

In the experimental $^2$H~NMR literature data used in this work~\cite{browning80,roux90},
the lipids were first solvated to the buffer and then centrifuged to a pellet that
was used in the measurements. Such samples have a lower lipid concentration
(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
Larger multilamellar repeat distances are expected in the samples with lower lipid
concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
However, the PS headgroup order parameters measured from gravimetric samples in this work
are in good agreement with the results from centrifuged samples in the literature~\cite{browning80} (Fig.~\ref{HGorderParameters}).
Furthermore, the equilibrium repeat distance rapidly decreases with the addition of monovalent
salts, and is close to the typical simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
in the used simulations and reference experiments.

Two different definitions for the salt concentrations have been used when the molecular
electrometer concept is applied to study ion binding affinities.
The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. The choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(section \ref{concentrationDEFsection} in the supplementary information).
